Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-07 paper2.qmd
section:all; process:write,figures
section:all; process:write,figures,code

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates  [@Perron_et_al_2015, pp. 2]. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b, pp. 267]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b, pp. 266]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp. 452]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 3. But are our specific problems? -->
However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use its measurement scores for hypothesis testing. Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677] (see @tbl-thurstone_cases). Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269] and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not address how to use the resulting measurement scores for hypothesis testing. Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp. 677]. Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
As previously discussed, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp. 16; @McElreath_2021]. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- 2. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Thus, given that modern CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice. These assumptions are the equal dispersion and zero correlation between stimuli.

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, the thought experiment occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], in this case, the assumption of equal dispersions. For instance, when the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts. This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data. In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376; @Bramley_2008, pp. 273; @Kelly_et_al_2022, pp. 678], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp. 11; @Pollitt_2012b, pp. 290]. These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, comparing misfit texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them.

@@

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Furthermore, if researchers acknowledge that misfit statistics may represent outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

<!-- 3. But does it hold?  -->
Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp. 268]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, when assessed, are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp. 346]. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 and @vanDaal_et_al_2016 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 3. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010, pp. 86]. For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or general quality. Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.

@@

<!-- 3. But what are the challenges when this assumption is violated? -->
Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues. Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023, pp. 341, 482]. These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020, pp. 370] and, by extension, the underlying discriminal processes of the stimuli. This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.

@@

Finally, as discussed in the previous section, removing *misfit* judges based risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis. *Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a, pp. 164-165; @Pollitt_2012b, pp. 289-290; @vanDaal_et_al_2016, pp. 4; @Goossens_et_al_2018, pp. 20].

@@

@@

<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp. 269]. Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests. For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error $(\sigma_{T})$ by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.

@@

@@

<!-- 4. segway for next section -->
In aggregate, the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates. This overreliance may also undermine their validity [@Perron_et_al_2015, pp. 2], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes. However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp. 4]. *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 4. But why identification is so important? -->
However, @sec-theory-issues presents compelling evidence that Thurstone's Case V, and by extension the BTL model, suffers from several statistical and measurement limitations. These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the example. Identification is crucial because it is a necessary condition for ensuring consistent estimators. *Consistency* refers to the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]. Without identification, consistency cannot be achieved, even with "infinite" and error-free data. Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023, pp. 5].

@@

<!-- 5. So, how do SCMs and DAGs support identification? -->
Fortunately, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [see @Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis. Second, they depict causal relationships in a non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp. 35].

@@

::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
$$

@@

Conceptual-population model, vectorized form.
:::

@@

<!-- Brought Rubin's missing data framework to the field. -->

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015]. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.

<!-- [@Perron_et_al_2015, pp. 2] -->

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

<!-- [@Thurstone_1927b, pp. 267] -->

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp. 452]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

<!-- [@Bramley_2008, pp. 251] -->

@@

<!-- 3. But are our specific problems? -->
However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use its measurement scores for hypothesis testing. Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases). Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269] and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not address how to use the resulting measurement scores for hypothesis testing. Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022]. Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?

<!-- [@Bramley_2008, pp. 253] -->

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
As previously discussed, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases). However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021]. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

<!-- [@Bramley_2008, pp. 254] -->

<!-- [@vanderLinden_et_al_2017_I, pp. 16] -->

@@

<!-- 2. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a]. Thus, given that modern CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice. These assumptions are the equal dispersion and zero correlation between stimuli.

<!-- [@Thurstone_1927a, pp. 376] -->

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, the thought experiment occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a]. Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019], in this case, the assumption of equal dispersions. For instance, when the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts. This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data. In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b]. These misfit texts may exhibit larger discriminal dispersions or represent outliers --texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, comparing misfit texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them.

@@

<!-- [@Pollitt_2004, pp. 11]  -->
<!-- [@Pollitt_2012b, pp. 290] -->

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Furthermore, if researchers acknowledge that misfit statistics may represent outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

<!-- 3. But does it hold?  -->
Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.

<!-- [@Thurstone_1927b, pp. 268] -->

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, when assessed, are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II]. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 and @vanDaal_et_al_2016 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 3. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010]. For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or general quality. Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.

<!-- [@Everitt_et_al_2010, pp. 86] -->

@@

<!-- 3. But what are the challenges when this assumption is violated? -->
Thus, erroneously assuming zero correlation between stimuli can also lead to significant statistical and measurement issues. Specifically, neglecting judges' biases or relevant hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Ackerman_1989; @Hoyle_et_al_2023]. These inaccuracies can result in spurious conclusions about the discriminal differences [@McElreath_2020] and, by extension, the underlying discriminal processes of the stimuli. This issue is illustrated in @fig-correlation when the discriminal difference distribution of the BTL scores follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $\rho \neq 0$.

@@

<!-- [@McElreath_2020, pp. 370] -->

Finally, as discussed in the previous section, removing *misfit* judges based risks discarding valuable information and even introduce bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable because they depend on which judges researchers exclude from the analysis. *Misfit judges* are those whose evaluations deviate substantially from the shared consensus due to the unique characteristics of either the stimuli or the judges themselves [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018].

@@

<!-- [@Pollitt_2012a, pp. 164-165]  -->
<!-- [@Pollitt_2012b, pp. 289-290] -->
<!-- [@vanDaal_et_al_2016, pp. 4] -->
<!-- [@Goossens_et_al_2018, pp. 20] -->

@@

<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
Building on the previous section, it is clear that, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b]. Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests. For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- [@Thurstone_1927b, pp. 269] -->

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023; @Hoyle_et_al_2023]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error $(\sigma_{T})$ by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.

@@

<!-- [@Kline_et_al_2023, pp. 25]  -->
<!-- [@Hoyle_et_al_2023, pp. 137] -->

@@

<!-- 4. segway for next section -->
In aggregate, the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates. This overreliance may also undermine their validity [@Perron_et_al_2015], particularly when coupled with the disconnect between the trait measurement and hypothesis testing processes. However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests. *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 4. But why identification is so important? -->
However, @sec-theory-issues presents compelling evidence that Thurstone's Case V, and by extension the BTL model, suffers from several statistical and measurement limitations. These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the example. Identification is crucial because it is a necessary condition for ensuring consistent estimators. *Consistency* refers to the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]. Without identification, consistency cannot be achieved, even with "infinite" and error-free data. Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].

@@

<!-- 5. So, how do SCMs and DAGs support identification? -->
Fortunately, SCMs and DAGs support identification analysis through two key advantages[^1]. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis. Second, they depict causal relationships in a non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016].

<!-- [@Pearl_et_al_2016, pp. 35] -->

[^1]: These topics are beyond the scope of this study, thus, readers seeking a more profound understanding can refer to introductory papers such as @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020, and introductory books like @Pearl_et_al_2018, @Neal_2020, and @McElreath_2020 are useful. For more advanced study, seminal papers such as @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, along with books such as @Pearl_2009, @Morgan_et_al_2014, and @Hernan_et_al_2020, are recommended.

@@

::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
$$

@@

Conceptual-population model, initial vectorized form.
:::

@@
